Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-17 cordillera-cycle.tex
Typeset: change fonts in math subscripts.

      A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
      relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the
      sliding velocity, $\vec{v}_b$ (Table~\ref{tab:params}). The yield
      stress, $\tau_c$, is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\degree as a~piecewise-linear function of
      modern bed elevation, with the lowest value occuring below modern sea
      level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
      value occuring above the generalised elevation of the highest
      shorelines \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus
      accounting for a~weakening of till associated with the presence of
      marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a~formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
      ratio and $C_c$ is a~measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{max}=2$\,\unit{m}, a~threshold above which additional
      melt water is assumed to drain off instantaneously. Finally, the
      bedrock topography responds to ice load following a~bedrock
      deformation model that includes local isostasy, elastic lithosphere
      flexure and viscous astenosphere deformation in an infinite half-space
      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
      value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is used for the
      astenosphere (Table~\ref{tab:params}) in accordance with the results
      from regional glacial isostatic adjustment modelling at the northern
      Cascadia subduction zone \citep{James.etal.2009}.

@@

      Ice surface accumulation and ablation are computed from monthly mean
      near-surface air temperature, $T_m$, monthly standard deviation of
      near-surface air temperature, $\sigma$, and monthly precipitation,
      $P_m$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
      Accumulation is equal to precipitation when air temperatures are below
      0\,\unit{{\degree}C}, and decreases to zero linearly with temperatures
      between 0 and 2\,\unit{{\degree}C}. Ablation is computed from PDD,
      defined as an integral of temperatures above 0\,\unit{{\degree}C} in
      one year.

      The PDD computation accounts for stochastic temperature variations by
      assuming a~normal temperature distribution of standard deviation
      $\sigma$ aroung the expected value $T_m$. It is expressed by an
      error-function formulation \citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology,
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

@@

      Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
      palaeo-temperature proxy records from the Greenland Ice Core Project
      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the
      Vostok ice core \citep{Petit.etal.1999}, and Ocean Drilling Program
      (ODP) sites 1012 and 1020, both located off the coast of California
      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP
      and NGRIP records were calculated from oxygen isotope
      (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
    {\Delta}T_{TS}(t) = &-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).

      Finally, lapse-rate corrections, ${\Delta}T_{LR}$, are computed as
      a~function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{ref}$:
\begin{align}
    {\Delta}T_{LR}(t, x, y) &= -\gamma [s(t, x, y)-b_{ref}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{ref}],
\end{align}

      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
      one hand, and for differences between the basal topography of the ice
      flow model, $b$, and the NARR reference topography, $b_{ref}$, on the
      other hand. All simulations use an annual temperature lapse rate of
      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer
      to different model runs by the name of the proxy record used for the
      palaeo-temperature forcing.

@@

    $v_{th}$& Pseudo-plastic threshold velocity
            & 100.0
            & \unit{m\,yr^{-1}} \\

@@

    $C_c$   & Till compressibility coefficient
            & 0.12
            & - \\

    $W_{max}$ & Maximal till water thickness
            & 2.0
            & m \\

@@

    $\rho_b$& Bedrock density
            & 3300
            & \unit{kg\,m^{-3}} \\

    $c_b$   & Bedrock specific heat capacity
            & 1000
            & \unit{J\,kg^{-1}\,K^{-1}} \\

    $k_b$   & Bedrock thermal conductivity
            & 3.0
            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\

@@

    $T_s$   & Temperature of snow precipitation
            & 273.15
            & \unit{K} \\

    $T_r$   & Temperature of rain precipitation
            & 275.15
            & \unit{K} \\

    $F_s$   & Degree-day factor for snow
            & $3.04\times10^{-3}$
            & \unit{m\,K^{-1}\,day^{-1}} \\

    $F_i$   & Degree-day factor for ice
            & $4.59\times10^{-3}$
            & \unit{m\,K^{-1}\,day^{-1}} \\

@@

\begin{figure*}
  \includegraphics{hr_maps_deglac}
  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
           Dashed segments (A--D) indicate the location of profiles used in
           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
  \label{fig:hr_geom_deglacageshots}
\end{figure*}

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{hr_geom_deglacage}
  \caption{Modelled age of the last deglaciation. Areas that have been covered
           only before the last glacial maximum are marked
           in green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,\unit{ka}..
           Dashed segments (A--D) indicate the location of profiles used in
           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
  \label{fig:hr_geom_deglacage}
\end{figure*}

>>>>

      A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
\end{equation}
      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
      sliding velocity, $\vec{v}_{\mathrm{b}}$ (Table~\ref{tab:params}). The yield
      stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
\end{equation}
      where cohesion, $c_0$, is assumed to be zero. The friction angle,
      $\phi$, varies from 15 to 45\degree as a~piecewise-linear function of
      modern bed elevation, with the lowest value occuring below modern sea
      level (0\,\unit{m} above sea level, \unit{m\,a.s.l.}) and the highest
      value occuring above the generalised elevation of the highest
      shorelines \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}, thus
      accounting for a~weakening of till associated with the presence of
      marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}). Effective pressure, $N$, is related to
      the ice overburden stress, $\rho gh$, and the modelled amount of
      subglacial water, using a~formula derived from laboratory experiments
      with till extracted from the base of Ice Stream B in West Antarctica
      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
\end{equation}
      where $\delta$ is chosen as 0.02, $e_0$ is a~measured reference void
      ratio and $C_{\mathrm{c}}$ is a~measured compressibility coefficient
      (Table~\ref{tab:params}). The amount of water at the base, $W$, varies
      from zero to $W_{\text{max}}=2$\,\unit{m}, a~threshold above which additional
      melt water is assumed to drain off instantaneously. Finally, the
      bedrock topography responds to ice load following a~bedrock
      deformation model that includes local isostasy, elastic lithosphere
      flexure and viscous astenosphere deformation in an infinite half-space
      \citep{Lingle.Clark.1985,Bueler.etal.2007}. A~relatively low viscosity
      value of $\nu_{m} = 1\times10^{19}$\,\unit{Pa\,s} is used for the
      astenosphere (Table~\ref{tab:params}) in accordance with the results
      from regional glacial isostatic adjustment modelling at the northern
      Cascadia subduction zone \citep{James.etal.2009}.

@@

      Ice surface accumulation and ablation are computed from monthly mean
      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation of
      near-surface air temperature, $\sigma$, and monthly precipitation,
      $P_{\mathrm{m}}$, using a~temperature-index model \citep[e.g.,][]{Hock.2003}.
      Accumulation is equal to precipitation when air temperatures are below
      0\,\unit{{\degree}C}, and decreases to zero linearly with temperatures
      between 0 and 2\,\unit{{\degree}C}. Ablation is computed from PDD,
      defined as an integral of temperatures above 0\,\unit{{\degree}C} in
      one year.

      The PDD computation accounts for stochastic temperature variations by
      assuming a~normal temperature distribution of standard deviation
      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed by an
      error-function formulation \citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology,
$\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{\text{TS}}$, and lapse-rate corrections, ${\Delta}T_{\text{LR}}$:
\begin{align}
    T_{\mathrm{m}}(t, x, y) &= T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
    P_{\mathrm{m}}(t, x, y) &= P_{\mathrm{m}0}(x, y) \,.
\end{align}

@@

      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
      palaeo-temperature proxy records from the Greenland Ice Core Project
      \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice Core
      Project \citep[NGRIP,][]{Andersen.etal.2004}, the European Project for
      Ice Coring in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, the
      Vostok ice core \citep{Petit.etal.1999}, and Ocean Drilling Program
      (ODP) sites 1012 and 1020, both located off the coast of California
      \citep{Herbert.etal.2001}. Palaeo-temperature anomalies from the GRIP
      and NGRIP records were calculated from oxygen isotope
      (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
    {\Delta}T_{\text{TS}}(t) = &-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. All records were scaled linearly
      (Table~\ref{tab:records}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).

      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed as
      a~function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{\text{ref}}$:
\begin{align}
    {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
\end{align}

      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
      one hand, and for differences between the basal topography of the ice
      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
      other hand. All simulations use an annual temperature lapse rate of
      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer
      to different model runs by the name of the proxy record used for the
      palaeo-temperature forcing.

@@

    $v_{\text{th}}$& Pseudo-plastic threshold velocity
            & 100.0
            & \unit{m\,yr^{-1}} \\

@@

    $C_{\mathrm{c}}$   & Till compressibility coefficient
            & 0.12
            & - \\

    $W_{\text{max}}$ & Maximal till water thickness
            & 2.0
            & m \\

@@

    $\rho_{\mathrm{b}}$& Bedrock density
            & 3300
            & \unit{kg\,m^{-3}} \\

    $c_{\mathrm{b}}$   & Bedrock specific heat capacity
            & 1000
            & \unit{J\,kg^{-1}\,K^{-1}} \\

    $k_{\mathrm{b}}$   & Bedrock thermal conductivity
            & 3.0
            & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}} \\

@@

    $T_{\mathrm{s}}$   & Temperature of snow precipitation
            & 273.15
            & \unit{K} \\

    $T_{\mathrm{r}}$   & Temperature of rain precipitation
            & 275.15
            & \unit{K} \\

    $F_{\mathrm{s}}$   & Degree-day factor for snow
            & $3.04\times10^{-3}$
            & \unit{m\,K^{-1}\,day^{-1}} \\

    $F_{\mathrm{i}}$   & Degree-day factor for ice
            & $4.59\times10^{-3}$
            & \unit{m\,K^{-1}\,day^{-1}} \\

@@

\begin{figure*}
  \includegraphics{hr_maps_deglac}
  \caption{Snapshots of modelled surface topography (200\,\unit{m} contours)
           and surface velocity (colour mapping) during the last deglaciation
           from the GRIP (top panels) and EPICA (bottom panels) 5\,\unit{km} simulations.
           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
  \label{fig:hr_geom_deglacageshots}
\end{figure*}

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{hr_geom_deglacage}
  \caption{Modelled age of the last deglaciation. Areas that have been covered
           only before the last glacial maximum are marked
           in green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,\unit{ka}..
           Dashed segments \textbf{(A--D)} indicate the location of profiles used in
           Figs.~\ref{fig:hr_pf_grip} and~\ref{fig:hr_pf_epica}.}
  \label{fig:hr_geom_deglacage}
\end{figure*}
